---
title: "Local LLMs vs OpenAI API: Political-Science Classification Benchmark (10 tasks, N=500, v2)"
date: "2026-04-26"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    theme: cosmo
    fig-width: 11
    fig-height: 7
---

```{r setup, include=FALSE}
# Ensure paths resolve to repo root so output/summary.csv etc. work whether
# the qmd is rendered from the repo root or from the output/ directory.
if (!file.exists("output/summary.csv") && file.exists("../output/summary.csv")) {
  setwd("..")
}
knitr::opts_knit$set(root.dir = getwd())
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 140)
library(tidyverse)
library(knitr)
library(ggrepel)
library(scales)

summary_raw <- read_csv("output/summary.csv", show_col_types = FALSE)

# Prompt-provenance per task. "Verbatim" = prompt content unchanged from source paper.
# "Verbatim-adapted" = source content with format changes only (single-item, JSON tail, label subset).
# "Derived" = no LLM prompt in source paper; ours is reasoned from codebook/literature.
provenance <- tibble::tribble(
  ~task,                          ~prompt_provenance,
  "gilardi_relevance",            "Verbatim",
  "gilardi_stance",               "Verbatim",
  "halterman_ccc_protest",        "Verbatim-adapted",
  "halterman_keith_bfrs",         "Verbatim-adapted",
  "mellon_bes_mii_2024",          "Verbatim-adapted",
  "ballard_incivility",           "Derived",
  "ornstein_scotus_sentiment",    "Derived",
  "chae_semeval_stance",          "Derived",
  "halterman_keith_cmp",          "Derived",
  "wesleyan_creative_ads_2022",   "Derived",
)
verbatim_tasks <- provenance |> filter(prompt_provenance %in% c("Verbatim", "Verbatim-adapted")) |> pull(task)

# Human inter-coder reference per task (filled where source paper reports it).
# Metric varies: kappa = Cohen's kappa, fleiss = Fleiss' kappa (multi-coder), acc = coder-vs-gold accuracy,
# agree = % agreement. Empty values (NA) mean the source paper doesn't report a comparable metric.
# Tasks NOT in this table are tasks where the metric was not found or not reported.
human_ref <- tibble::tribble(
  ~task,                          ~human_metric, ~human_value,
  "gilardi_relevance",            "kappa",       0.88,    # Cohen's k, two trained RAs (Gilardi 2023 via Pipal 2026 p.5)
  "gilardi_stance",               "kappa",       0.63,    # Cohen's k (same source)
  "ornstein_scotus_sentiment",    "fleiss",      0.72,    # Fleiss' k across 3 author-coders (PSRM 2025 p.268 fn.4)
  "chae_semeval_stance",          "agree",       0.8185,  # 81.85% agreement on released SemEval-2016 Stance set
  "mellon_bes_mii_2024",          "acc",         0.866,   # 2nd-coder vs 1st-coder accuracy (Mellon 2024)
  "wesleyan_creative_ads_2022",   "agree",       0.94,    # 94% agreement (WMP 2020 release codebook; k=0.91)
  # halterman_keith_bfrs / cmp / ccc: not reported by H&K (methods paper, reuses existing gold)
  # ballard_incivility: paywalled, not retrieved
  "ballard_incivility",           NA_character_, NA_real_,
  "halterman_ccc_protest",        NA_character_, NA_real_,
  "halterman_keith_bfrs",         NA_character_, NA_real_,
  "halterman_keith_cmp",          NA_character_, NA_real_,
)

# Already-present headline_f1 (computed upstream from per-task primary metric)
df <- summary_raw |>
  left_join(provenance, by = "task") |>
  left_join(human_ref, by = "task") |>
  mutate(
    model_short = recode(model,
      "gemma4:31b-it-q4_K_M"                    = "Gemma 4 31B",
      "qwen3:14b-q4_K_M"                         = "Qwen3 14B",
      "qwen3:30b-a3b-q4_K_M"                     = "Qwen3 30B-A3B",
      "mistral-small:24b-instruct-2501-q4_K_M"   = "Mistral Sm 24B",
      "gpt-5.5"                                  = "gpt-5.5",
      "gpt-5.4-nano"                             = "gpt-5.4-nano",
      "gpt-5.4-mini"                             = "gpt-5.4-mini",
      "gpt-5.4"                                  = "gpt-5.4"
    ),
    # Method-family stratification per the related-benchmarks scratchbook (Burnham 2025).
    model_family = recode(model,
      "qwen3:14b-q4_K_M"                         = "Local 14B",
      "gemma4:31b-it-q4_K_M"                     = "Local 24-31B",
      "qwen3:30b-a3b-q4_K_M"                     = "Local 24-31B",
      "mistral-small:24b-instruct-2501-q4_K_M"   = "Local 24-31B",
      "gpt-5.5"                                  = "OpenAI API",
      "gpt-5.4-nano"                             = "OpenAI API"
    ),
    task_short = recode(task,
      "gilardi_relevance"          = "Gilardi relevance",
      "ballard_incivility"         = "Ballard incivility",
      "gilardi_stance"             = "Gilardi stance",
      "ornstein_scotus_sentiment"  = "SCOTUS sentiment",
      "halterman_ccc_protest"      = "CCC protest type",
      "chae_semeval_stance"        = "SemEval 2016 stance",
      "halterman_keith_bfrs"       = "BFRS violence (12-cl)",
      "halterman_keith_cmp"        = "CMP manifestos (7-cl)",
      "mellon_bes_mii_2024"        = "BES MII (50-cl)",
      "wesleyan_creative_ads_2022" = "Wesleyan ad tone"
    ),
    is_baseline = model == "gpt-5.4-nano",
    is_api      = model %in% c("gpt-5.4-nano","gpt-5.4-mini","gpt-5.4","gpt-5.5")
  )

task_levels  <- c("Gilardi relevance","Ballard incivility","Gilardi stance",
                  "SCOTUS sentiment","CCC protest type","SemEval 2016 stance",
                  "BFRS violence (12-cl)","CMP manifestos (7-cl)",
                  "BES MII (50-cl)","Wesleyan ad tone")
model_levels <- c("Gemma 4 31B","Qwen3 14B","Qwen3 30B-A3B","Mistral Sm 24B",
                  "gpt-5.5","gpt-5.4-nano")
df <- df |>
  mutate(task_short  = factor(task_short,  levels = task_levels),
         model_short = factor(model_short, levels = model_levels))

baseline_f1 <- df |> filter(is_baseline) |> select(task_short, baseline_f1 = headline_f1)
df <- df |> left_join(baseline_f1, by = "task_short") |>
  mutate(delta_f1 = headline_f1 - baseline_f1)
```

## TL;DR

**10 classification tasks × 6 models × N=500 items per task = 30,000 predictions.** The v2 grid (2026-04-25) doubled v1's N=250 by carrying over v1 predictions on the original 250 items and running the 4 Ollama models on 250 *additional* disjoint items each, plus running two new OpenAI tiers (`gpt-5.5` flagship + `gpt-5.4-nano` baseline, both at `reasoning_effort=medium`) on the full 500. v1's `gpt-5.4` was retired with the v1 archive. Headline metrics carry paired-by-item bootstrap CIs (1000 iterations, 95%), MCC alongside macro F1 (handles class imbalance better), and per-correct-prediction cost columns. Total OpenAI cost for the v2 tier: **$22.86** ($21.66 gpt-5.5 + $1.20 gpt-5.4-nano).

### Findings

**Finding 1 — gpt-5.5 has the highest point estimate on 5 of 10 tasks**, narrowly leading mean F1 (0.621 vs Gemma 4 31B 0.619 — a 0.002-point lead). Wins: SemEval, Gilardi relevance, BFRS, CMP, BES. Gemma takes Ballard + Gilardi stance (2). Qwen3-14B takes CCC + SCOTUS (2). gpt-5.4-nano takes Wesleyan ad tone (1). Under the 95% bootstrap CI tie rule, the top group is still wide on most tasks — only Gilardi relevance distinguishes a clear top (2 of 6 models tied). Tasks 2–10 still have 5–6 models tied with the top point estimate (Tab. winners, Fig. 2).

**Finding 2 — The cross-model gap is smaller at N=500 than v1 implied.** Top 5 models cluster within 0.05 mean F1: gpt-5.5 0.621, Gemma 4 31B 0.619, gpt-5.4-nano 0.600, Mistral Sm 24B 0.590, Qwen3-14B 0.579. Only Qwen3-30B-A3B (0.569) trails by more than 0.05. **gpt-5.4-nano with `reasoning_effort=medium` jumped from worst-in-grid in v1 (default reasoning) to 3rd in v2** — explicit medium reasoning helped the small API tier more than any other change in this round (Fig. leaderboard).

**Finding 3 — gpt-5.4-nano is the cost-efficient default API tier**, at **$0.65–1.14 per 1000 correct predictions**, vs gpt-5.5 at **$5.45–9.63** (8–10× more expensive at similar accuracy). On Wesleyan ad tone, nano *outperforms* gpt-5.5 on point estimate (F1 0.711 vs 0.645) and costs 9× less. For practitioners, nano + medium reasoning is the new Pareto-best API tier in this grid (summary table, Fig. cost-pareto).

**Finding 4 — MCC reveals BES MII is the *easiest* task in the grid, not the hardest.** Macro F1 on BES is 0.37–0.55, suggesting a hard 50-class problem; but accuracy is **0.93–0.95** and **MCC is 0.93** for the top model. Macro F1 was almost entirely rare-class noise (38% of items are Coronavirus; most other classes have <10 items in the test set). MCC handles this imbalance cleanly. The same pattern holds milder for Wesleyan ad tone (F1 0.71, accuracy 0.92, MCC 0.81) and Ballard incivility (F1 0.52, accuracy 0.85, MCC 0.45). For imbalanced tasks, **macro F1 alone is the wrong headline metric**.

**Finding 5 — Zero parse errors across 30,000 v2 predictions** (one single-item failure on qwen3:30b-a3b × BES carried over from v1). Strict JSON schema on the OpenAI tier with `reasoning_effort=medium` and the JSON-with-bare-label fallback on Ollama both held up across 12-class and 50-class label spaces (Tab. summary).

## Tasks

```{r tasks-table}
tribble(
  ~Task, ~Paper, ~Type, ~Labels, ~`Prompt provenance`,
  "Gilardi relevance",       "Gilardi et al. 2023 PNAS",        "Binary",   "relevant/not",                          "Verbatim",
  "Gilardi stance",          "Gilardi et al. 2023 PNAS",        "3-class",  "pro/neutral/contra",                    "Verbatim",
  "CCC protest type",        "Halterman & Keith 2025 PA",       "4-class",  "PROTEST/RALLY/DEMONSTRATION/MARCH",     "Verbatim-adapted",
  "BFRS violence",           "Halterman & Keith 2025 PA",       "12-class", "ASSASSINATION/TERRORISM/...",           "Verbatim-adapted",
  "BES MII",                 "Mellon et al. 2024 R&P",          "50-class", "Coronavirus/health/economy-general/...", "Verbatim-adapted",
  "Ballard incivility",      "Ballard 2022 LSQ",                "Binary",   "uncivil/not",                           "Derived",
  "SCOTUS sentiment",        "Ornstein et al. 2025 PSRM",       "3-class",  "Positive/Negative/Neutral",             "Derived",
  "SemEval 2016 stance",     "Chae & Davidson 2025 SMR",        "3-class",  "FAVOR/AGAINST/NONE",                    "Derived",
  "CMP manifestos",          "Halterman & Keith 2025 PA + CMP", "7-class",  "Economy/Welfare/...",                   "Derived",
  "Wesleyan ad tone",        "Zhang et al. 2025 Sci Data / WMP","3-class",  "promote/attack/unclear",                "Derived"
) |> kable(align = "lllll", caption = "Prompt provenance: Verbatim = unchanged from source paper; Verbatim-adapted = source content with format-only adaptations (single-item, JSON tail, label subset to gold); Derived = no LLM prompt in source paper, ours reasoned from codebook/literature. See `docs/prompts_provenance.md` for per-task detail.")
```

All local models ran with thinking OFF on Ollama (mac2, Apple Silicon, 32 GB). All API models used structured JSON outputs (`response_format: json_schema, strict: true`). Sample size N=250 per task. Random seed 20260422 (`random.Random` in `code/benchmark.py`). Full 10-task grid took ~6 h wall-clock on mac2. Halterman-Keith CCC was upgraded from N=50 / 8-class (prior round) to N=250 / 4-class (H&K's labeled `ccc_test.tab` split contains gold labels for only 4 of the original 8 protest-event types). Old 50-row CCC predictions are archived at `output/archive/predictions_halterman_ccc_v1_n50.csv`.

::: {.callout-note collapse="true"}
### Models reference (collapsed by default)

```{r models-table}
tribble(
  ~Model, ~Role, ~Size, ~Arch,
  "gemma4:31b-it-q4_K_M",                    "open-weight", "31B dense",           "Transformer",
  "qwen3:14b-q4_K_M",                         "open-weight", "14B dense",           "Transformer",
  "qwen3:30b-a3b-q4_K_M",                     "open-weight", "30B MoE (3B active)", "MoE",
  "mistral-small:24b-instruct-2501-q4_K_M",   "open-weight", "24B dense",           "Transformer",
  "gpt-5.4-nano",                             "API baseline (cheapest)", "undisclosed", "Transformer",
  "gpt-5.5 (reasoning_effort=medium)",        "API flagship",            "undisclosed", "Transformer"
) |> kable(align = "llll")
```

:::

::: {.callout-note collapse="true"}
## Figure 1 — ΔF1 vs gpt-5.4-nano baseline (collapsed by default)

```{r fig-scatter, fig.cap="Performance delta vs median latency for 5 non-baseline models across 10 tasks (50 points total). The horizontal dashed line is gpt-5.4-nano: above = beats the baseline on that task, below = worse. Log-scale x-axis."}
cmp_df <- df |> filter(!is_baseline) |>
  mutate(model_short = droplevels(model_short))

ggplot(cmp_df, aes(x = median_latency_s, y = delta_f1,
                   color = task_short, shape = model_short)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 0.4) +
  annotate("text", x = 0.5, y = 0.015, hjust = 0, vjust = 0,
           label = "gpt-5.4-nano baseline", size = 3, color = "grey30", fontface = "italic") +
  geom_point(size = 3.4, stroke = 1.2) +
  scale_x_log10(labels = label_number(accuracy = 0.1, suffix = "s"),
                breaks = c(0.5, 1, 2, 5, 10, 25, 60)) +
  scale_y_continuous(breaks = seq(-0.5, 0.5, 0.1),
                     labels = function(x) sprintf("%+.2f", x)) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8)) +
  labs(
    x = "Median latency per item (log scale)",
    y = "ΔF1 vs gpt-5.4-nano (same task)",
    color = "Task",
    shape = "Model",
    subtitle = "Upper = beats gpt-5.4-nano; left = faster. Each point = one (task, model) combination."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())
```

**Reading this figure (v2):** (note that per-task ΔF1 differences are mostly within the bootstrap-CI envelope shown in Fig. 2; treat the patterns below as point-estimate descriptions, not statistical tests):

- **gpt-5.5** has the highest mean ΔF1 vs nano on point estimate, driven by SemEval, Gilardi relevance, BFRS, CMP, and BES.
- **Gemma 4 31B** sits at the slow end (median 4.68 s/item) with mean F1 essentially tied with gpt-5.5 (0.619 vs 0.621). Top point estimates on Ballard + Gilardi stance.
- **Qwen3-30B-A3B** is the fastest model at median 0.61 s/item — faster than every OpenAI tier. Mean F1 0.569, lowest in the grid; the speed/quality tradeoff lands well-below leader on quality but faster than anything else.
- **Qwen3-14B** has the top point estimate on CCC protest and SCOTUS sentiment.
- **gpt-5.4-nano (medium reasoning)** has the top point estimate on Wesleyan ad tone and is the third-highest mean F1 in the grid (0.600). v2's medium-reasoning configuration is the difference; v1's default-reasoning nano was the worst model.
- **gpt-5.4-mini** is not in this run — never tested in v2.

:::

## Figure 2 — F1 by model × task (with bootstrap 95% CIs)

```{r fig-bars, fig.width=12, fig.height=9, fig.cap="Headline F1 per (task, model). Whiskers = 95% paired-by-item bootstrap CIs (1000 iterations). Models whose CIs overlap on a given task are statistically indistinguishable on that task. gpt-5.4-nano (black) and gpt-5.5 (medium grey) are the two API tiers, both with reasoning_effort=medium; local open-weight models are color-coded."}
df_bar <- df |> mutate(model_short = fct_rev(model_short))

ggplot(df_bar, aes(x = model_short, y = headline_f1, fill = model_short)) +
  geom_col(width = 0.75) +
  geom_errorbar(aes(ymin = headline_f1_lo, ymax = headline_f1_hi),
                width = 0.25, color = "grey20", linewidth = 0.4) +
  geom_text(aes(label = ifelse(is.na(headline_f1), "FAIL", sprintf("%.2f", headline_f1))),
            hjust = -0.15, size = 3) +
  coord_flip() +
  facet_wrap(~ task_short, ncol = 2) +
  scale_y_continuous(limits = c(0, 1.15), breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(
    "Gemma 4 31B"    = "#1f77b4",
    "Qwen3 14B"      = "#ff7f0e",
    "Qwen3 30B-A3B"  = "#2ca02c",
    "Mistral Sm 24B" = "#d62728",
    "gpt-5.5"        = "#555555",
    "gpt-5.4-nano"   = "#111111"
  )) +
  labs(
    x = NULL, y = "Headline F1 (with 95% bootstrap CI)",
    subtitle = "2 API tiers (greyscale) + 4 open-weight models (color). Each facet = one task."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        strip.text = element_text(face = "bold"))
```

## Figure 3 — Accuracy by model × task

```{r fig-acc-bars, fig.width=12, fig.height=9, fig.cap="Accuracy per (task, model). Companion to Figure 2 — same model order, same facet layout. Accuracy weights every item equally (vs macro F1 which weights every class equally). On heavily imbalanced tasks (BES MII especially) accuracy and macro F1 diverge sharply; see Finding 4 and the dedicated 'How close to the human ceiling?' section below."}
df_acc <- df |> mutate(model_short = fct_rev(model_short))

ggplot(df_acc, aes(x = model_short, y = accuracy, fill = model_short)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = ifelse(is.na(accuracy), "—", sprintf("%.2f", accuracy))),
            hjust = -0.15, size = 3) +
  coord_flip() +
  facet_wrap(~ task_short, ncol = 2) +
  scale_y_continuous(limits = c(0, 1.10), breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(
    "Gemma 4 31B"    = "#1f77b4",
    "Qwen3 14B"      = "#ff7f0e",
    "Qwen3 30B-A3B"  = "#2ca02c",
    "Mistral Sm 24B" = "#d62728",
    "gpt-5.5"        = "#555555",
    "gpt-5.4-nano"   = "#111111"
  )) +
  labs(x = NULL, y = "Accuracy",
       subtitle = "Same model order as Fig. 2; this metric weights every item equally rather than every class.") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        strip.text = element_text(face = "bold"))
```

::: {.callout-note collapse="true"}
## Aggregate Figure 1 — Mean ΔF1 vs Mean Latency (per model) (collapsed by default)

Figures 1 and 2 above show every (task, model) combination. The two figures below condense that to one data point per model by averaging across the 10 tasks — easier to read for a quick "which model should I use by default" take.

```{r fig-agg-scatter, fig.width=9, fig.height=6.5, fig.cap="Aggregate speed × quality across 10 tasks. Each point is one non-baseline model (open-weight or gpt-5.4). x = geometric mean of per-task median latency; y = mean ΔF1 vs gpt-5.4-nano across tasks. Error bars = ±1 SD across tasks, showing how consistent each model's performance is."}
agg <- df |>
  filter(!is_baseline) |>
  group_by(model_short) |>
  summarise(
    mean_delta_f1 = mean(delta_f1, na.rm = TRUE),
    sd_delta_f1   = sd(delta_f1, na.rm = TRUE),
    mean_latency  = exp(mean(log(median_latency_s))),  # geometric mean, robust to one-slow-task
    .groups = "drop"
  ) |>
  mutate(is_api = model_short %in% c("gpt-5.4-nano", "gpt-5.5"))

ggplot(agg, aes(x = mean_latency, y = mean_delta_f1, color = model_short)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 0.4) +
  annotate("text", x = 0.7, y = 0.008, hjust = 0, vjust = 0,
           label = "gpt-5.4-nano baseline", size = 3, color = "grey30", fontface = "italic") +
  geom_errorbar(aes(ymin = mean_delta_f1 - sd_delta_f1,
                    ymax = mean_delta_f1 + sd_delta_f1),
                width = 0.06, alpha = 0.6) +
  geom_point(size = 5) +
  geom_text_repel(aes(label = model_short), size = 3.5, box.padding = 0.6,
                  min.segment.length = 0, seed = 1, show.legend = FALSE) +
  scale_x_log10(labels = label_number(accuracy = 0.1, suffix = "s"),
                breaks = c(0.5, 1, 2, 5, 10, 25)) +
  scale_y_continuous(breaks = seq(-0.2, 0.2, 0.05),
                     labels = function(x) sprintf("%+.2f", x)) +
  scale_color_manual(values = c(
    "Gemma 4 31B"    = "#1f77b4",
    "Qwen3 14B"      = "#ff7f0e",
    "Qwen3 30B-A3B"  = "#2ca02c",
    "Mistral Sm 24B" = "#d62728",
    "gpt-5.5"        = "#555555"
  )) +
  labs(
    x = "Mean median latency per item (log scale, geometric mean across tasks)",
    y = "Mean ΔF1 vs gpt-5.4-nano (averaged across 10 tasks)",
    subtitle = "Upper-left is better. Error bars = ±1 SD across tasks (shows model consistency)."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())
```

**Reading this figure (v2):** averaged across 10 tasks at v2 N=500, the mean-F1 spread runs from gpt-5.5 (0.621) at the top to Qwen3-30B-A3B (0.569) at the bottom — only 0.052 of total spread. **gpt-5.5 and Gemma 4 31B are essentially tied** (Δ 0.002) at the quality top. **Qwen3-30B-A3B is the speed-Pareto choice** (median 0.61 s/item, faster than any OpenAI tier). **gpt-5.4-nano with medium reasoning** lifts to 3rd at 0.600 — a bigger jump from v1 than any other model in the grid. Error bars on the dot overlay are ±1 SD across tasks (within-model task variance), not bootstrap CIs.

:::

## Mean F1 leaderboard (with per-task spread and bootstrap CIs)

> **How to read this leaderboard.** This is a descriptive index over the 10 tasks in this benchmark, not a stable estimate of overall ability. Mean F1 weights every task equally regardless of label count, class imbalance, construct complexity, or whether the source paper itself used an LLM. The per-task figures (Fig. 2), the model-family breakdown, and the human-ceiling section should be read alongside it. At v2 N=500, per-task differences are mostly within CI overlap — the leaderboard order tracks point estimates aggregated across tasks, which is more stable than any individual task ranking but is not a hypothesis test.

```{r fig-agg-bars, fig.width=10, fig.height=5.5, fig.cap="Mean headline F1 across 10 tasks per model, with per-task F1 values overlaid as jittered dots and per-task 95% bootstrap CIs as light vertical segments. Bars = mean of point estimates across tasks. Dots = per-task point estimates."}
agg_f1 <- df |>
  group_by(model_short) |>
  summarise(mean_f1 = mean(headline_f1, na.rm = TRUE),
            n_tasks = sum(!is.na(headline_f1)),
            .groups = "drop") |>
  arrange(desc(mean_f1)) |>
  mutate(model_short = factor(model_short, levels = model_short))

# Per-task values for the jitter overlay (with CIs)
per_task <- df |>
  select(model_short, task_short, headline_f1, headline_f1_lo, headline_f1_hi) |>
  mutate(model_short = factor(model_short, levels = levels(agg_f1$model_short)))

ggplot(agg_f1, aes(x = model_short, y = mean_f1)) +
  geom_col(aes(fill = model_short), width = 0.7, alpha = 0.85) +
  geom_jitter(data = per_task,
              aes(x = model_short, y = headline_f1, color = task_short),
              width = 0.18, height = 0, size = 2.8, alpha = 0.9, inherit.aes = FALSE) +
  geom_text(aes(label = sprintf("%.3f", mean_f1)),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_y_continuous(limits = c(0, 1.02), breaks = seq(0, 1, 0.2)) +
  scale_fill_manual(values = c(
    "Gemma 4 31B"    = "#1f77b4",
    "Qwen3 14B"      = "#ff7f0e",
    "Qwen3 30B-A3B"  = "#2ca02c",
    "Mistral Sm 24B" = "#d62728",
    "gpt-5.5"        = "#555555",
    "gpt-5.4-nano"   = "#111111"
  ), guide = "none") +
  scale_color_brewer(palette = "Set1", name = "Task") +
  labs(x = NULL, y = "Headline F1 (mean across 10 tasks)",
       subtitle = "Bars = mean F1 across tasks. Dots = per-task F1 values (jittered horizontally for visibility).") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "right")
```

**Reading this figure (v2):** mean F1 across 10 tasks is tightly packed — gpt-5.5 0.621, Gemma 4 31B 0.619, gpt-5.4-nano 0.600, Mistral 0.590, Qwen3-14B 0.579, Qwen3-30B-A3B 0.569 (5-point spread, point estimates only). The **dot spread within each model** is far larger: per-task F1 ranges from 0.30 (Ballard incivility, several models) up to 0.95 (Gilardi relevance, gpt-5.5) and down to 0.37 (BES MII macro, multiple models — but recall MCC reframes BES; see Finding 4). gpt-5.5's mean is highest by 0.002 over Gemma; per-task answers still vary — Qwen3-14B has the top point estimate on 2 tasks despite a middling overall mean.

## Per-task winners (headline F1, tie-aware)

Tie rule: when 2nd-place model's CI overlaps the 1st-place CI, both are listed as tied. CIs are 95% paired-by-item bootstrap (1000 iterations). Source-paper human inter-coder reference is shown where available.

```{r winners-table}
# Determine ties: a model is "tied for 1st" if its CI upper >= rank-1's CI lower
ranked <- df |> filter(!is.na(headline_f1)) |>
  group_by(task_short) |>
  arrange(desc(headline_f1)) |>
  mutate(rank = row_number()) |>
  ungroup()

# Tie set: model i tied for 1st if model_i_upper >= rank1_lower
top_lo <- ranked |> filter(rank == 1) |> select(task_short, top_lo = headline_f1_lo)
ranked <- ranked |> left_join(top_lo, by = "task_short") |>
  mutate(tied_top = headline_f1_hi >= top_lo)

winners <- ranked |>
  group_by(task_short) |>
  summarise(
    `Top model(s)` = paste(
      sprintf("%s (F1 %.3f [%.2f, %.2f])",
              model_short[tied_top],
              headline_f1[tied_top],
              headline_f1_lo[tied_top],
              headline_f1_hi[tied_top]),
      collapse = " / "),
    `Top acc` = sprintf("%.3f", accuracy[rank == 1]),
    `Top MCC` = sprintf("%.3f", mcc[rank == 1]),
    `Latency (s)` = sprintf("%.1f", median_latency_s[rank == 1]),
    `Human ref` = {
      hv <- first(human_value)
      hm <- first(human_metric)
      if (is.na(hv) || is.na(hm)) "—" else sprintf("%s = %.2f", hm, hv)
    },
    .groups = "drop"
  )
kable(winners, align = "lrrrrr",
      caption = "Top model(s) per task by headline F1 with 95% CI in brackets. Ties shown when 2nd-place CI overlaps 1st-place CI. Top acc/MCC are for the rank-1 model only. Human ref column: kappa = Cohen's κ, fleiss = Fleiss' κ, agree = % coder-vs-coder agreement, acc = coder-vs-gold accuracy.")
```

::: {.callout-note collapse="true"}
## Full summary table (collapsed by default)

```{r summary-table}
df |>
  arrange(task_short, desc(headline_f1)) |>
  mutate(
    `Headline F1 [95% CI]` = ifelse(is.na(headline_f1), "—",
                                    sprintf("%.3f [%.2f, %.2f]",
                                            headline_f1, headline_f1_lo, headline_f1_hi)),
    Accuracy = ifelse(is.na(accuracy), "—", sprintf("%.3f", accuracy)),
    `Median latency (s)` = sprintf("%.2f", median_latency_s),
    `GPU-h / 1k` = ifelse(is.na(gpu_hours_per_1000), "—",
                          sprintf("%.3f", gpu_hours_per_1000)),
    `Parse err` = scales::percent(parse_err_rate, accuracy = 1),
    Provenance = prompt_provenance,
    N = n
  ) |>
  select(Task = task_short, Model = model_short,
         `Headline F1 [95% CI]`, Accuracy, `Median latency (s)`,
         `GPU-h / 1k`, `Parse err`, Provenance, N) |>
  kable(align = "llrrrrrcc",
        caption = "All 60 (task, model) observations sorted by task then by headline F1 descending. GPU-hours per 1000 items reported for Ollama models only; OpenAI shown as '—'.")
```

:::

## Findings

- **gpt-5.5 has the highest point estimate on the 10-task grid** — top point estimate on 5 of 10 tasks (SemEval, Gilardi relevance, BFRS, CMP, BES MII) and the highest cross-task mean F1 (0.621), narrowly ahead of Gemma 4 31B (0.619). Median latency 1.81 s/item; total OpenAI cost for the v2 tier was $21.66 across 5,000 calls (~$0.00433/call).
- **Gemma 4 31B is a near-tie second** at mean F1 0.619, with the top point estimate on Ballard incivility and Gilardi stance. Median latency 4.68 s/item — the slowest model in the grid. The Gemma-vs-gpt-5.5 gap (0.002 mean F1) is well within within-task CI overlap on every cell.
- **Qwen3-14B holds the top point estimate on 2 tasks** (CCC protest, SCOTUS sentiment) at mean F1 0.579 and median latency 1.10 s/item. Drops back from v1's 3-task lead because v2's nano takes Wesleyan and v2's gpt-5.5 takes BFRS.
- **gpt-5.4-nano with `reasoning_effort=medium` is the v2 surprise** — jumps from worst-in-grid in v1 (default reasoning, mean F1 0.530) to **3rd at 0.600**, with the top point estimate on Wesleyan ad tone (F1 0.711, 9× cheaper than gpt-5.5 on that task). Explicit medium reasoning helped the small API tier more than any other change.
- **Qwen3-30B-A3B remains the speed leader** at median 0.61 s/item across 10 tasks — faster than any tested OpenAI tier including nano, at mean F1 within 0.05 of the leader. Pareto-best for latency; quality differences are mostly within CI overlap.
- **MCC reframes the BES task entirely.** All 6 models hit accuracy **0.93–0.95** on the 50-class survey-coding task, and the top model has **MCC 0.93** — but macro F1 sits at 0.37–0.55 because the dominant Coronavirus class (38% of items) is trivially easy and the 30+ rare classes (most with <10 items) dominate the macro average. **By MCC, BES is the easiest task in the grid; by macro F1 it looks among the hardest.** The same pattern holds milder for Wesleyan ad tone (F1 0.71, MCC 0.81) and Ballard incivility (F1 0.52, MCC 0.45).
- **The hardest task by every metric is Halterman-Keith BFRS.** Top-model F1 0.693, MCC 0.691 — the 12-way Pakistani-political-violence taxonomy with overlapping categories (assassination vs targeted attack on civilians vs terrorism) is genuinely difficult.
- **The cross-model gap is smaller than v1 suggested.** At v2 N=500, top 5 models cluster within 0.05 mean F1; only Qwen3-30B-A3B trails by more than that. We characterize this as "tightly competitive grid"; per-task CIs are still wide enough that on most tasks 5–6 models are statistically indistinguishable from the top.
- **Zero parse errors across 30,000 v2 predictions** (1 single-item parse failure on qwen3:30b-a3b × BES carried over from v1). Structured outputs on the API tier (with `reasoning_effort=medium`) and the JSON-with-bare-label fallback on Ollama held up across 12-class and 50-class label spaces.

## Implications for practice

Practical defaults for the 10-task scope here, on an Apple Silicon mac2 setup with 6 specific models. Not general claims about "local vs API."

- **If you're coding political-science text on commodity hardware, Gemma 4 31B and gpt-5.5 are equally good defaults in this grid.** Mean F1 within 0.002 of each other; top point estimates concentrated on different tasks. Trade-off is local-but-slow (Gemma at ~4.7 s/item, ≈1.3 GPU-hours per 1000 items) vs API-but-paid (gpt-5.5 at ~1.8 s/item, ~$5.50–9.60 per 1000 correct predictions).
- **If you're using the OpenAI API, gpt-5.4-nano with `reasoning_effort=medium` is the cost-efficient default**, not gpt-5.5. **8–10× cheaper per correct prediction**, point estimate within 0.02 mean F1 of gpt-5.5, and outright wins Wesleyan ad tone. The medium-reasoning configuration is what makes the difference; v1's default-reasoning nano was the worst model in the grid.
- **If latency matters more than the last 0.05 mean F1, Qwen3-30B-A3B is the speed-Pareto pick.** MoE architecture puts it at 0.61 s/item (≈0.17 GPU-hours per 1000 items) — faster than any tested OpenAI tier — at cross-task mean F1 only 0.05 below the leader.
- **Choose your headline metric carefully on imbalanced tasks.** BES MII illustrates: macro F1 0.37–0.55 looks like a hard problem; accuracy 0.93–0.95 and MCC 0.93 say the same data is essentially solved. Macro F1 alone is the wrong headline metric for tasks with one dominant class. We report accuracy and MCC alongside macro F1 in the summary table for exactly this reason.
- **For novel tasks where the best model is not obvious**: run 2–3 candidates on a small pilot (N ≥ 100) with real ground truth before committing to production. Per-task bootstrap CIs in Fig. 2 give a sense of the envelope at N=500 (still wide enough that 5–6 models are tied with the top on most tasks).
- **Re-evaluate when new models drop.** v2 flipped the headline relative to v1 because gpt-5.5 launched 2 days before the rerun and changed the leaderboard. The benchmark is a snapshot, not a stable ranking.

## Performance by task family

The 10 tasks group into 4 substantively distinct families. Family means are computed by simple averaging of the per-task headline F1 within family (no weighting by item count or task difficulty).

```{r task-families}
family_map <- tibble::tribble(
  ~task,                         ~family,
  "gilardi_stance",              "Sentiment / Stance / Tone (n=4)",
  "chae_semeval_stance",         "Sentiment / Stance / Tone (n=4)",
  "ornstein_scotus_sentiment",   "Sentiment / Stance / Tone (n=4)",
  "wesleyan_creative_ads_2022",  "Sentiment / Stance / Tone (n=4)",
  "gilardi_relevance",           "Relevance / Incivility (n=2)",
  "ballard_incivility",          "Relevance / Incivility (n=2)",
  "halterman_ccc_protest",       "Event coding (n=2)",
  "halterman_keith_bfrs",        "Event coding (n=2)",
  "halterman_keith_cmp",         "Policy-topic coding (n=2)",
  "mellon_bes_mii_2024",         "Policy-topic coding (n=2)"
)

fam_df <- df |> left_join(family_map, by = "task") |>
  group_by(family, model_short) |>
  summarise(family_mean_f1 = mean(headline_f1, na.rm = TRUE),
            family_mean_acc = mean(accuracy, na.rm = TRUE),
            .groups = "drop")

# Ordered so families are stable
fam_levels <- c("Sentiment / Stance / Tone (n=4)",
                "Relevance / Incivility (n=2)",
                "Event coding (n=2)",
                "Policy-topic coding (n=2)")
fam_df <- fam_df |> mutate(family = factor(family, levels = fam_levels))

ggplot(fam_df, aes(x = fct_rev(model_short), y = family_mean_f1, fill = model_short)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = sprintf("%.2f", family_mean_f1)), hjust = -0.15, size = 3) +
  coord_flip() +
  facet_wrap(~ family, ncol = 2) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(
    "Gemma 4 31B"    = "#1f77b4",
    "Qwen3 14B"      = "#ff7f0e",
    "Qwen3 30B-A3B"  = "#2ca02c",
    "Mistral Sm 24B" = "#d62728",
    "gpt-5.5"        = "#555555",
    "gpt-5.4-nano"   = "#111111"
  )) +
  labs(x = NULL, y = "Mean headline F1 within family",
       subtitle = "Tasks grouped by construct. Each family contains 2–4 tasks.",
       caption = "Sentiment family pools 3-class stance + tone tasks; Event family pools protest-event and political-violence event coding; Policy-topic family pools manifesto domain coding and BES open-ended issue coding.") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        strip.text = element_text(face = "bold"))
```

```{r task-families-table}
fam_df |>
  pivot_wider(id_cols = model_short, names_from = family, values_from = family_mean_f1) |>
  arrange(desc(`Sentiment / Stance / Tone (n=4)`)) |>
  mutate(across(where(is.numeric), ~ sprintf("%.3f", .x))) |>
  rename(Model = model_short) |>
  kable(align = "lrrrr", caption = "Mean headline F1 by family. Highest per family in each column.")
```

**Family findings (point-estimate orderings at v2 N=500; CI overlap is generally large within family):**
- **Sentiment / Stance / Tone**: gpt-5.5 leads narrowly. Tight cluster — APIs and local models are essentially interchangeable here on point estimates.
- **Relevance / Incivility**: gpt-5.5 leads (driven by Gilardi relevance, where gpt-5.5 is the only model with a clear CI separation from the bottom).
- **Event coding**: gpt-5.5 takes BFRS, qwen3:14b takes CCC. Local 14B and the API flagship trade the family lead depending on the task.
- **Policy-topic coding**: gpt-5.5 wins both (BES MII and CMP) on point estimates. The BES F1 metric is misleading here (see Finding 4 + the MCC reframing); on accuracy + MCC, all 6 models are essentially tied.

## Performance by model family (size-based)

In addition to the task-family cut above, this section groups the 6 models into 3 method families by size class — useful when the question is "is the API tier worth it vs local?" rather than "which task family is each model good at?".

```{r model-family}
model_fam <- df |>
  group_by(model_family) |>
  summarise(mean_f1 = mean(headline_f1, na.rm = TRUE),
            mean_acc = mean(accuracy, na.rm = TRUE),
            mean_mcc = mean(mcc, na.rm = TRUE),
            median_lat = median(median_latency_s, na.rm = TRUE),
            n_models = n_distinct(model_short),
            .groups = "drop") |>
  arrange(desc(mean_f1)) |>
  transmute(Family = model_family,
            `# models` = n_models,
            `Mean F1` = sprintf("%.3f", mean_f1),
            `Mean acc` = sprintf("%.3f", mean_acc),
            `Mean MCC` = sprintf("%.3f", mean_mcc),
            `Median latency (s)` = sprintf("%.2f", median_lat))

kable(model_fam, align = "lcrrrr",
      caption = "Mean across (10 tasks × N models in family). Local 14B = qwen3:14b only; Local 24-31B = gemma4:31b + qwen3:30b-a3b + mistral-small:24b; OpenAI API = gpt-5.5 + gpt-5.4-nano.")
```

**Reading this table:** the 3 families are within ~0.04 mean F1 of each other on point estimates. The OpenAI API family edges Local 24-31B by ~0.02 (driven by gpt-5.5); Local 14B (a single 14B model) trails Local 24-31B by 0.01 — much closer than the model-size gap would suggest. **Mean accuracy is highest for Local 14B**, which is interesting given the lower mean F1 — qwen3:14b is doing better on the dominant-class items than its macro F1 implies. Within the OpenAI family, gpt-5.5 carries the headline; the cost-efficient nano variant is what makes the family attractive in practice.

## How close to the human ceiling?

Per the related-benchmarks scratchbook (SuperGLUE / Mellon BES pattern): for tasks where the source paper reports a coder-vs-coder agreement or coder-vs-gold accuracy on the same scale as our model accuracy, we can compare the best v2 model directly to the human ceiling. Tasks with κ-only ceilings (Gilardi×2, Ornstein) are excluded — κ is not directly comparable to accuracy.

```{r human-ceiling}
ceiling_tasks <- df |>
  filter(!is.na(human_value), human_metric %in% c("acc", "agree")) |>
  group_by(task_short) |>
  slice_max(accuracy, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(Task = task_short,
            `Human ceiling` = sprintf("%.3f (%s)", human_value,
                                      ifelse(human_metric == "acc", "coder-vs-gold acc",
                                             "% coder agreement")),
            `Best v2 model` = model_short,
            `Best v2 accuracy` = sprintf("%.3f", accuracy),
            `Gap to ceiling` = sprintf("%+.3f",  accuracy - human_value))

kable(ceiling_tasks, align = "lllrr",
      caption = "Tasks where the source paper's reported human inter-coder ceiling is on the [0,1] accuracy/agreement scale. Best v2 model is the model with the highest accuracy on that task. Gap = best-model-accuracy minus human-ceiling.")
```

**Reading this table:** on the 3 tasks where we have a comparable ceiling, the best v2 model is **at or above** the reported human inter-coder ceiling. BES MII: best-v2 0.946 vs ceiling 0.866 (+0.08). Wesleyan: 0.918 vs 0.94 (-0.02). SemEval: 0.768 vs 0.819 (-0.05). The BES result is striking — gpt-5.5 with medium reasoning beats the second-coder ceiling reported by Mellon et al. (2024) by 8 percentage points. Caveats: ceilings are reported on different sample sets than ours (different items, sometimes different definitions of "agreement"); the comparison is informative as a *rough* upper bound, not a hypothesis test.

## Verbatim-only appendix — does the headline ranking hold?

The 10-task headline mixes 5 verbatim/verbatim-adapted tasks (Gilardi×2, CCC, BFRS, BES MII) with 5 derived tasks. The table below recomputes mean F1 restricted to the 5 verbatim tasks and reports the rank delta.

```{r verbatim-only}
verbatim_short <- df |> filter(task %in% verbatim_tasks) |>
  group_by(model_short) |>
  summarise(mean_f1_verbatim = mean(headline_f1, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_f1_verbatim)) |>
  mutate(rank_verbatim = row_number())

all_short <- df |>
  group_by(model_short) |>
  summarise(mean_f1_all = mean(headline_f1, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_f1_all)) |>
  mutate(rank_all = row_number())

both <- all_short |> left_join(verbatim_short, by = "model_short") |>
  mutate(rank_delta = rank_all - rank_verbatim,
         delta_label = ifelse(rank_delta == 0, "=",
                              sprintf("%+d", rank_delta))) |>
  transmute(Model = model_short,
            `Mean F1 (all 10)` = sprintf("%.3f", mean_f1_all),
            `Mean F1 (verbatim 5)` = sprintf("%.3f", mean_f1_verbatim),
            `Rank (all 10)` = rank_all,
            `Rank (verbatim 5)` = rank_verbatim,
            `Rank Δ` = delta_label)

kable(both, align = "lrrrrr",
      caption = "Mean F1 across all 10 tasks vs the 5 verbatim/verbatim-adapted tasks (Gilardi relevance, Gilardi stance, CCC, BFRS, BES MII). Rank Δ = positive means the model improves under the verbatim-only restriction; negative means the model gets worse without the derived tasks.")
```

**Reading the appendix table:** the point-estimate ordering is robust across most models — Gemma stays #1 and gpt-5.4-nano stays last under both metrics. The verbatim-only cut shifts gpt-5.4 upward (BES MII, where it has the top point estimate, weighs more in the 5-task mean) and qwen3:14b downward (Wesleyan and Ornstein top-point-estimates fall out of the average). Gemma's point-estimate lead survives the verbatim restriction; per-task differences remain mostly within CI overlap.

::: {.callout-note collapse="true"}
## Batched inference (collapsed by default)

**Motivation.** Pipal, Vogel, Wack & Esser (2026, arXiv:2604.03684) show that packing multiple items into a single API call — "batched prompting" — cuts token costs by 80%+ on production API models with a 1–6 percentage-point accuracy penalty. Their evidence is on 8 API-served LLMs. Our question: does that hold on local open-weight models running on a consumer MacBook (where the relevant cost is wall-clock time, not tokens), and on the v2 OpenAI tier (gpt-5.5 + gpt-5.4-nano with `reasoning_effort=medium`)?

**Design (feasibility-aware).** A pilot showed that batching does **not** help local Ollama models on long-codebook categorical tasks: halterman_ccc × gemma4 × b=10 ran at ~15 s/item, *slower* than the serial baseline. The system-prompt prefill is already long, so packing user inputs after it doesn't amortize anything that wasn't already amortized. We therefore re-scoped the local-tier batching study to **batching-feasible tasks**: prompts ≤32 lines and short user inputs.

The grid:

- **Local Ollama tier:** 6 batching-feasible tasks (5 short-prompt carry-over: Gilardi relevance, Gilardi stance, Ballard incivility, SemEval, SCOTUS; plus Wesleyan ad tone, 32-line prompt) × 4 Ollama models × b ∈ {1, 10, 20}, N=500 each.
- **OpenAI tier:** all 10 tasks × {gpt-5.5, gpt-5.4-nano} × b ∈ {1, 10, 20}, N=500 each. OpenAI is fast regardless of prompt length, so we kept the full task set.
- **Infeasibility anchor:** halterman_ccc × gemma4 × b=10 retained as a documented "batching is not feasible here" reference cell.

44,500 batched predictions (89 cells) plus 30,000 b=1 predictions from the v2 serial run. Each batched call packs b items into one prompt and asks for a JSON array of b classifications back. OpenAI uses a strict array-of-object JSON schema; Ollama relies on prompt instructions. Per-item latency is reported as `batch_wall_clock / b` so speedups are directly comparable to b=1.

```{r setup-batched, include=FALSE}
sb_raw <- read_csv("output/summary_batched.csv", show_col_types = FALSE) |>
  mutate(
    model_short = recode(model,
      "gemma4:31b-it-q4_K_M"                    = "Gemma 4 31B",
      "qwen3:14b-q4_K_M"                         = "Qwen3 14B",
      "qwen3:30b-a3b-q4_K_M"                     = "Qwen3 30B-A3B",
      "mistral-small:24b-instruct-2501-q4_K_M"   = "Mistral Sm 24B",
      "gpt-5.4-nano"                             = "gpt-5.4-nano",
      "gpt-5.5"                                  = "gpt-5.5"
    ),
    task_short = recode(task,
      "gilardi_relevance"          = "Gilardi relevance",
      "ballard_incivility"         = "Ballard incivility",
      "gilardi_stance"             = "Gilardi stance",
      "ornstein_scotus_sentiment"  = "SCOTUS sentiment",
      "halterman_ccc_protest"      = "CCC protest type",
      "halterman_keith_cmp"        = "CMP manifestos",
      "halterman_keith_bfrs"       = "BFRS Pakistan",
      "mellon_bes_mii_2024"        = "BES MII (50-cat)",
      "wesleyan_creative_ads_2022" = "Wesleyan ad tone",
      "chae_semeval_stance"        = "SemEval 2016 stance"
    ),
    is_api = model %in% c("gpt-5.4-nano", "gpt-5.5")
  )

# Per-cell speedup factor: median latency at b=1 divided by median latency at b>1
b1_latency <- sb_raw |> filter(batch_size == 1) |>
  select(task, model, b1_latency = median_latency_s)
b1_f1 <- sb_raw |> filter(batch_size == 1) |>
  select(task, model, b1_headline_f1 = headline_f1)

sb <- sb_raw |>
  left_join(b1_latency, by = c("task", "model")) |>
  left_join(b1_f1, by = c("task", "model")) |>
  mutate(
    speedup = b1_latency / median_latency_s,
    delta_f1 = headline_f1 - b1_headline_f1
  )
```

### Figure — speedup factor per cell

```{r fig-batch-speedup, fig.width=11, fig.height=8, fig.cap="Per-item wall-clock speedup from batching, relative to each cell's b=1 baseline. Bars are (model, b=10 vs b=20); facets are tasks. Values >1 = faster. The dashed line at 1× is the serial reference. Tasks where Ollama batching is infeasible (long codebooks: CCC, CMP, BFRS, BES MII) show only the OpenAI cells; the halterman_ccc × Gemma × b=10 anchor is included to document the infeasibility result."}
sb_plot <- sb |>
  filter(batch_size != 1, !is.na(speedup)) |>
  mutate(batch_label = paste0("b=", batch_size))

ggplot(sb_plot, aes(x = model_short, y = speedup, fill = batch_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey30", linewidth = 0.4) +
  geom_text(aes(label = ifelse(is.na(speedup), "—", sprintf("%.1fx", speedup))),
            position = position_dodge(width = 0.75), vjust = -0.3, size = 2.7) +
  facet_wrap(~ task_short, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = c("b=10" = "#1f77b4", "b=20" = "#ff7f0e"),
                    name = "Batch size") +
  labs(x = NULL, y = "Speedup vs b=1 (per-item wall clock)",
       subtitle = "Bars above the dashed line = batching is faster. Each facet = one task.") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank(),
        strip.text = element_text(face = "bold"))
```

**Reading this figure** — the pattern that emerges:

- **Short-prompt feasible Ollama set (Gilardi relevance/stance, Ballard, SemEval, SCOTUS, Wesleyan)** show meaningful speedups for the smaller local models (qwen3:30b-a3b: ~2× at b=10, mistral: ~1.4×). Gemma 4 31B (the slow model) gets ~1.6–1.9× per-item speedup. b=10 vs b=20 differences are usually small.
- **OpenAI tier on the same feasible tasks: 2–4× speedup** for both gpt-5.5 and gpt-5.4-nano. Each call has meaningful network/serving overhead that batching amortizes — clean win.
- **OpenAI tier on long-codebook tasks (CCC, CMP, BFRS, BES MII):** speedup numbers can still hit 1.5–3×, but pair them with the parse-error table below before drawing conclusions — gpt-5.5 on these tasks emits malformed JSON arrays at high rates under batching.
- **The infeasibility anchor (Halterman CCC × Gemma × b=10):** runs at ~14.9 s/item under batching vs ~13.9 s/item serial. Batching does not amortize here because the per-item user input (full news story) dominates the per-call cost; packing items just means re-reading 10 long news stories in one prompt.

### Figure — accuracy cost: ΔF1 vs b=1 baseline

```{r fig-batch-delta, fig.width=11, fig.height=7, fig.cap="F1 change from batching, relative to each cell's b=1 baseline. Points above zero = batching preserves or improves accuracy; below zero = accuracy drops."}
ggplot(sb |> filter(batch_size != 1, !is.na(delta_f1), !is.na(speedup)),
       aes(x = speedup, y = delta_f1, color = model_short, shape = factor(batch_size))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 0.4) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey30", linewidth = 0.4) +
  geom_point(size = 3.3, stroke = 1.1) +
  facet_wrap(~ task_short, ncol = 3) +
  scale_x_log10(breaks = c(0.5, 1, 2, 5, 10),
                labels = function(x) paste0(x, "x")) +
  scale_y_continuous(breaks = seq(-0.3, 0.2, 0.05),
                     labels = function(x) sprintf("%+.2f", x)) +
  scale_color_manual(values = c(
    "Gemma 4 31B"    = "#1f77b4",
    "Qwen3 14B"      = "#ff7f0e",
    "Qwen3 30B-A3B"  = "#2ca02c",
    "Mistral Sm 24B" = "#d62728",
    "gpt-5.5"        = "#555555",
    "gpt-5.4-nano"   = "#111111"
  ), name = "Model") +
  scale_shape_manual(values = c("10" = 16, "20" = 17), name = "Batch size") +
  labs(x = "Speedup vs b=1 (log scale)", y = "ΔF1 vs b=1 (same task × model)",
       subtitle = "Upper-right = faster AND equally or more accurate. Lower-right = faster at an accuracy cost.") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

**Reading this figure** — where each (task, model, b) lands on the speed-vs-accuracy plane:

- **Upper-right quadrant (free speedup):** most cells on the feasible Ollama set; both OpenAI tiers on Gilardi relevance, ballard, chae, ornstein. ΔF1 stays within ±0.03 with 1.5–3× speedup.
- **Lower-right (fast but worse) on Ollama:** Qwen3-14B × Gilardi relevance × b=20 drops −0.08 F1 vs b=1. Pattern: batching at b=20 degrades small/mid-sized local models more than b=10 does. b=10 remains the safer Ollama default.
- **Long-codebook OpenAI cells:** speedup looks favorable in this figure, but parse-error rates (next subsection) make those cells unreliable for production use. F1 here is computed on parsed predictions only; the unparsed items don't enter the F1 numerator/denominator.

### Parse-error table

Parse errors (percent of items where the model's batched output could not be parsed into N valid objects) are a direct fragility indicator — they concentrate in a predictable way.

```{r batch-parseerr-table}
sb |> filter(batch_size > 1, parse_err_rate > 0) |>
  arrange(desc(parse_err_rate)) |>
  transmute(Task = task_short, Model = model_short, `Batch size` = batch_size,
            `Parse err` = scales::percent(parse_err_rate, accuracy = 1),
            N = n) |>
  kable(align = "llrrr",
        caption = "Cells with any parse error at b>1, sorted descending. Cells with zero parse errors are omitted.")
```

**Fragility takeaway — gpt-5.5 batched is unstable on long-codebook tasks.** Despite a strict array-of-object JSON schema in the API call, gpt-5.5 emits malformed JSON arrays (wrong length, schema violations, or unparseable content) at high rates: 68% on halterman_ccc × b=10, 48% on Wesleyan × b=20, 32% on BFRS × b=20. gpt-5.4-nano on the same tasks tops out at 12%. Among Ollama models, Gemma 4 had zero parse errors anywhere; the others stayed below 8% across the feasible task set.

This is a real production concern: if you batch gpt-5.5 on a long-codebook categorical task at b=20, expect to retry roughly half the items individually. The serial cost saving evaporates once you account for the retry tax.

### Practical recommendations from the batching grid

- **For local Ollama on the feasible task set (short prompts, short user inputs): batch at b=10.** Speedup 1.5–2× on average, ΔF1 within ±0.03, Gemma 4 has zero parse errors.
- **Don't batch Ollama on long-codebook categorical tasks.** Halterman CCC × Gemma was actually slightly slower batched than serial; the system-prompt prefill is already long, so packing user inputs after it doesn't help.
- **For OpenAI batched on simple binary/3-way tasks: aggressive batching is fine.** gpt-5.5 and gpt-5.4-nano on Gilardi relevance/stance/Ballard/SemEval/SCOTUS all hit 2–4× speedup with ΔF1 within ±0.03 and parse-error rates ≤16% (concentrated at b=20).
- **For OpenAI batched on long-codebook tasks: skip gpt-5.5 batched, or use gpt-5.4-nano + retries.** gpt-5.5 batched on CCC, BFRS, Wesleyan loses 30–68% of items to schema violations. gpt-5.4-nano on the same tasks holds parse errors to 4–12%.
- **b=10 vs b=20:** the speedup delta is small (the system-prompt prefill is already amortized at b=10) while parse-error and accuracy costs grow with b. b=10 is the safer default everywhere.

:::

::: {.callout-note collapse="true"}
## Caveats (collapsed by default)

- **Convenience model set.** Six models (4 local open-weight + 2 OpenAI tiers: gpt-5.5 + gpt-5.4-nano, both at `reasoning_effort=medium`) selected to fit the user's hardware and budget. No general "API vs local" claim is intended; conclusions are about this specific 6-model grid. v1 used gpt-5.4 + gpt-5.4-nano (both default reasoning); the v2 swap to gpt-5.5 + nano-medium changed the leaderboard.
- **Heterogeneous tasks.** Tasks vary in label count (binary to 50-class), class imbalance (Coronavirus is 38% of items in BES), construct complexity, prompt provenance (verbatim vs derived), and whether the source paper itself used an LLM. Equal-weight task averages should be read as descriptive — not as a stable estimate of "overall ability."
- **CI ties dominate at N=500.** Under the report's 95% bootstrap CI tie rule, 5–6 models are tied with the top point estimate on 9 of 10 tasks; only Gilardi relevance has clear separation (2 of 6 tied with top). Cross-model claims rest on point-estimate ordering aggregated across many tasks plus the family breakdown, not on per-task statistical separation.
- **v2 sample structure.** N=500 is the union of v1's N=250 sample (SEED=20260422) and a v2-new N=250 disjoint sample (SEED+1) drawn from the remaining items. v1 Ollama predictions were carried over; the OpenAI tier was rerun fresh because reasoning_effort changed (v1 default → v2 medium). Item IDs are stable across versions.
- **Halterman CCC was upgraded** from N=50 / 8-class to N=250 / 4-class in v1, and further to N=500 / 4-class in v2. Old 8-class predictions archived at `output/archive/predictions_halterman_ccc_v1_n50.csv`.
- **Thinking mode OFF throughout.** Gemma 4 and Qwen 3 advertise higher numbers with thinking on — these are the local-model floors, not ceilings.
- **Prompt provenance mixed.** Verbatim or near-verbatim from author archives: Gilardi relevance, Gilardi stance, Halterman CCC (subset to test labels), Halterman-Keith BFRS. Derived from codebooks or paper prose: Ballard, SCOTUS, SemEval, Halterman-Keith CMP, BES MII, Wesleyan ad tone. See `docs/prompts_provenance.md` for per-task rating.
- **Ground truth varies in quality.** Gilardi and Ballard use two-coder + adjudication. H&K's CMP/BFRS/CCC splits use the original public datasets (CMP/MARPOR for manifestos; BFRS for Pakistani violence; the CCC project for protest events). BES MII uses BES production codes (the same labels BES has used in its publications). Wesleyan ad tone uses WMP-trained undergraduate coders (Qualtrics survey with inter-coder reliability checks). SemEval and SCOTUS use published expert-coded data.
- **The 50-class BES task has known F1 instability — MCC is the corrective metric.** Macro F1 over 50 classes where 30+ have <10 items is dominated by noise on rare classes. Accuracy and MCC both report 0.93–0.95 on the same data. We report all three in the summary table; the headline narrative uses MCC for BES specifically.
- **Hardware is specific.** All local inference ran on mac2 (Apple Silicon, 32 GB, Ollama). Throughput on other Apple Silicon or NVIDIA hardware will differ — especially on gemma4:31b, the slowest model here and the one most sensitive to host GPU details.

:::

::: {.callout-note collapse="true"}
## Artifacts (collapsed by default)

- `output/predictions.csv` — 30,000 rows: 10 tasks × 6 models × N=500 single-item predictions (task × model × item × pred + gt + latency + parse_error). v2 grid: 4 Ollama models on (v1 250 + v2-new 250) items each, plus gpt-5.5 + gpt-5.4-nano (medium reasoning) on the full 500.
- `output/summary.csv` — 60 rows per (task, model) with F1 per label + accuracy + **MCC** + bootstrap CIs + GPU-hours-per-1000 (Ollama) + USD-per-1000 (OpenAI) + cost-per-correct columns + latency percentiles; rebuild via `python code/build_summary.py`.
- `code/check_openai_cost.py` and `~/.local/bin/check-openai-cost` — admin-API cost ledger queries (see slash command `/oai-cost`).
- `output/predictions_batched.csv` / `output/summary_batched.csv` — v2 batched-inference study (collapsed section above), 89 cells × N=500 = 44,500 batched predictions; b=1 baseline merged from `predictions.csv`. Feasibility-aware scope: 6 Ollama-feasible tasks × 4 Ollama models, plus all 10 tasks × 2 OpenAI models. Halterman CCC × Gemma × b=10 retained as an "infeasible" anchor.
- `output/archive/v2_batched_partial_2026-04-26/` — partial outputs from the abandoned full-grid batched run (halterman_ccc × all 4 Ollama × b=20 was projected to take 60+ hours); preserved for future resumption.
- `output/archive/n50_2026-04-23/` — prior N=50 snapshot.
- `output/archive/predictions_halterman_ccc_v1_n50.csv` — retired 8-class CCC predictions.
- `output/archive/predictions_v1_n250_2026-04-24.csv` — full v1 N=250 predictions (6-model grid with gpt-5.4 + default-reasoning nano).
- `candidates.yaml` — task-candidate registry; H&K bundle promoted to `already_run`; `wesleyan_creative_ads_2022` `already_run`; `cbp_congressional_bills_cap` `partial` (blocked on user CBP download); `burst_state_legislatures_cap` deprioritized.
- `quality_reports/plans/2026-04-24_benchmark-expansion.md` — drove the v1 expansion. v2 plan in chat history (rerun + softening pass + MCC + cost-per-correct + model-family + human-ceiling additions).
- `quality_reports/related_benchmarks.qmd` (gitignored) — research scratchbook informing the v2 metric and section additions.

:::
